---
title: "Final Project"
subtitle: "ENVIRON 872"
author: "Flannery McNair, Andi Scroggs, and Alonso Valasquez"
date: "2025-12-05"
output: html_document
toc: TRUE
lof: TRUE
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE, include=FALSE}

library(here)
library(lubridate)
library(tidyverse)
library(readr)
library(dplyr)

here()
```



## Rationale and Research Question


\newpage

## Dataset Information


\newpage

## Exploration of Raw Data

Exploration of the raw data consisted of understanding the number of files and contents of those files for each Station. Originally, the water quality data and the nutrient data were separate for each station we are studying. We uploaded all of the files with nutrient and water quality data from the four stations we are studying for the years 2002-2024. We obtained a separate file with information about the stations in which the data was collected. 

```{r import of raw data, message=FALSE, warning=FALSE, include=FALSE}
sampling_stations <- read_csv("sampling_stations.csv")

# Importing the data within each folder as one and then combining data from each folder into their own datasets since they have the same columns

# Nutrients
nocecnut <- list.files(path = "./Data/nocecnut", pattern = "\\.csv$", full.names = TRUE)
nocec_nutrients <- nocecnut %>%
  plyr::ldply(read.csv)

noclcnut <- list.files(path = "./Data/noclcnut", pattern = "\\.csv$", full.names = TRUE)
noclc_nutrients <- noclcnut %>%
  plyr::ldply(read.csv)

nocrcnut <- list.files(path = "./Data/nocrcnut", pattern = "\\.csv$", full.names = TRUE)
nocrc_nutrients <- nocrcnut %>%
  plyr::ldply(read.csv)

noczbnut <- list.files(path = "./Data/noczbnut", pattern = "\\.csv$", full.names = TRUE)
noczb_nutrients <- noczbnut %>%
  plyr::ldply(read.csv)

# Water Quality 
nocecwq <- list.files(path = "./Data/nocecwq", pattern = "\\.csv$", full.names = TRUE)
nocec_water_quality <- nocecwq %>%
  plyr::ldply(read.csv)

noclcwq <- list.files(path = "./Data/noclcwq", pattern = "\\.csv$", full.names = TRUE)
noclc_water_quality <- noclcwq %>%
  plyr::ldply(read.csv)

nocrcwq <- list.files(path = "./Data/nocrcwq", pattern = "\\.csv$", full.names = TRUE)
nocrc_water_quality <- nocrcwq %>%
  plyr::ldply(read.csv)

noczbwq <- list.files(path = "./Data/noczbwq", pattern = "\\.csv$", full.names = TRUE)
noczb_water_quality <- noczbwq %>%
  plyr::ldply(read.csv)
```


\newpage

## Data Wrangling

After the conclusion of exploring the raw data, we began to wrangle the data by choosing the variables we wished to keep for our study. For the nutrient data sets we kept:
* Station Code
* Date and Time
* Orthophosphate (PO4F) (mg/L) 
* Ammonnium Fluoride (NH4F) (mg/L) 
* Nitryl Fluoride (NO2F) (mg/L) 
* Fluorine Nitrate (NO3F) (mg/L) 
* Nitrate + Nitrate (NO23F) (mg/L)
* Chlorophyll (ug/L)
For the water quality data sets we kept the variables:  
* Station Code
* Date and Time
* Temperature (degrees Celsius)
* Salinity (ppt)
* Dissolved Oxygen (%)
* Dissolved Oxygen in milligrams per Liter (mg/L)
* Depth (m)
* pH (standard units)
* Turbidity (NTU)

 
```{r first cleaning/choosing the columns, message=FALSE, warning=FALSE, include=FALSE}
# Nutrients
nocec_nutrients_cleaned <- nocec_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F, NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

noclc_nutrients_cleaned <- noclc_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F, NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

nocrc_nutrients_cleaned <- nocrc_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F,NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

noczb_nutrients_cleaned <- noczb_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F, NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

# Water Quality 
nocec_water_quality_cleaned <- nocec_water_quality %>%
  select(StationCode, DateTimeStamp, Temp, Sal, DO_Pct, 
         DO_mgl, Depth, pH, Turb)

noclc_water_quality_cleaned <- noclc_water_quality %>%
  select(StationCode, DateTimeStamp, Temp, Sal,  DO_Pct, 
         DO_mgl, Depth, pH, Turb)

nocrc_water_quality_cleaned <- nocrc_water_quality %>%
  select(StationCode, DateTimeStamp, Temp, Sal,  DO_Pct, 
         DO_mgl, Depth, pH, Turb)

noczb_water_quality_cleaned <- noczb_water_quality %>% 
  select(StationCode, DateTimeStamp, Temp, Sal,  DO_Pct, 
         DO_mgl, Depth, pH, Turb)
```

```{r combining the datasets, echo=FALSE, message=FALSE, warning=FALSE}
# Nutrients
nutrients <- bind_rows(nocec_nutrients_cleaned,
                       noclc_nutrients_cleaned,
                       nocrc_nutrients_cleaned,
                       noczb_nutrients_cleaned)
str(nutrients)

# Water Quality
water_quality <- bind_rows(nocec_water_quality_cleaned,
                           noclc_water_quality_cleaned,
                           nocrc_water_quality_cleaned,
                           noczb_water_quality_cleaned)
str(water_quality)
```

```{r creating a Date column, message=FALSE, warning=FALSE, include=FALSE}
# Nutrients
nutrients <- nutrients %>%
  mutate(DateTimeStamp = mdy_hm(DateTimeStamp))
  
nutrients <- nutrients %>% 
  mutate(Date = as.Date(DateTimeStamp))

# Water Quality
water_quality <- water_quality %>%
  mutate(DateTimeStamp = mdy_hm(DateTimeStamp))

water_quality <- water_quality %>%
  mutate(Date = as.Date(DateTimeStamp))
```

```{r clean Sampling Station dataset, echo=FALSE, message=FALSE, warning=FALSE}
# Sampling Stations
# rename the  columns bc they have spaces
sampling_stations <- sampling_stations %>%
  rename(NERR_SiteID = `NERR Site ID`,
    StationCode = `Station Code`,
    StationName = `Station Name`,
    LatLong = `Lat Long`,
    Reserve_Name = `Reserve Name`)

#Choosing the columns we want
sampling_stations_cleaned <- sampling_stations %>%
  select(NERR_SiteID, StationCode, StationName, LatLong, 
         Latitude, Longitude, Status, State, Reserve_Name)

sampling_stations_cleaned <- sampling_stations_cleaned %>%
  filter(StationCode %in% c("nocecnut", "nocecwq", "noclcnut", "noclcwq", 
                            "nocrcnut", "nocrcwq", "noczbnut", "noczbwq"))
str(sampling_stations_cleaned)
```

```{r clean up formatting of StationCodes}
# Clean up spaces and formatting differences
nutrients$StationCode <- trimws(nutrients$StationCode)
water_quality$StationCode <- trimws(water_quality$StationCode)
sampling_stations_cleaned$StationCode <- trimws(sampling_stations_cleaned$StationCode)

str(nutrients$StationCode)
str(water_quality$StationCode)
str(sampling_stations_cleaned$StationCode)
```

```{r Averaging the data per day, message=FALSE, warning=FALSE, include=FALSE}
# Nutrient - daily avg of all variables
nutrients_dailyavg <- nutrients %>%
  group_by(Date, StationCode) %>%
  summarise(PO4F = mean(PO4F, na.rm = TRUE),
            NH4F = mean(NH4F, na.rm = TRUE),
            NO2F = mean(NO2F, na.rm = TRUE),
            NO3F = mean(NO3F, na.rm = TRUE),
            NO23F = mean(NO23F, na.rm = TRUE),
            CHLA_N = mean(CHLA_N, na.rm = TRUE))

# Nutrient - monthly avg of all variables 
nutrients_monthlyavg <- nutrients %>%
  mutate(Year = year(Date),
         Month = month(Date)) %>%
  group_by(Year, Month, StationCode) %>%
  summarise(PO4F = mean(PO4F, na.rm = TRUE),
            NH4F = mean(NH4F, na.rm = TRUE),
            NO2F = mean(NO2F, na.rm = TRUE),
            NO3F = mean(NO3F, na.rm = TRUE),
            NO23F = mean(NO23F, na.rm = TRUE),
            CHLA_N = mean(CHLA_N, na.rm = TRUE))
  
# Water Quality - avg per day of all variables
water_quality_dailyavg <- water_quality %>%
  group_by(Date, StationCode) %>%
  summarise(Temp = mean(Temp, na.rm = TRUE),
            Sal = mean(Sal, na.rm = TRUE),
            DO_Pct = mean(DO_Pct, na.rm = TRUE),
            DO_mgl = mean(DO_mgl, na.rm = TRUE),
            Depth = mean(Depth, na.rm = TRUE),
            pH = mean(pH, na.rm = TRUE),
            Turb = mean(Turb, na.rm = TRUE))

# Water Quality - monthly avg of all variables 
water_quality_monthlyavg <- water_quality %>%
  mutate(Year = year(Date),
         Month = month(Date)) %>%
  group_by(Year, Month, StationCode) %>%
  summarise(Temp = mean(Temp, na.rm = TRUE),
            Sal = mean(Sal, na.rm = TRUE),
            DO_Pct = mean(DO_Pct, na.rm = TRUE),
            DO_mgl = mean(DO_mgl, na.rm = TRUE),
            Depth = mean(Depth, na.rm = TRUE),
            pH = mean(pH, na.rm = TRUE),
            Turb = mean(Turb, na.rm = TRUE))
```

After cleaning, preparing, and combining all of the files into one data set for nutrients and one data set for water quality, we added the sampling station data to each. This allowed us to create two large data sets which we will use for our study.

```{r add station names to nutrient and water quality data, message=FALSE, warning=FALSE, include=FALSE}
#joining station names to nutrients
nutrients_stations_dailyavg <- 
  left_join(nutrients_dailyavg, sampling_stations_cleaned, by = "StationCode")

nutrients_stations_monthlyavg <- 
  left_join(nutrients_dailyavg, sampling_stations_cleaned, by = "StationCode")

#joining station names to water_quality
water_quality_stations <- 
  left_join(water_quality_dailyavg, sampling_stations_cleaned, by = "StationCode")

water_quality_stations <- 
  left_join(water_quality_monthlyavg, sampling_stations_cleaned, by = "StationCode")
```

```{r}
#creating a proper date column for time series analysis
water_quality_stations <- water_quality_stations %>% 
  mutate(Date = ym(paste(Year, Month)))
```

Now we want to save this final cleaned data set and make sure it's pushed to our repository in Github!
```{r}
write.csv(water_quality_stations, row.names = FALSE, file =
            "./Data/water_quality_stations_cleaned")
```


\newpage

## Exploration of Processed Data

```{r}

```


\newpage

## Analysis


## Summary and Conclusion


\newpage

## References
