---
title: "Final Project"
subtitle: "ENVIRON 872"
author: "Flannery McNair, Andi Scroggs, and Alonso Valasquez"
date: "2025-12-05"
output: html_document
toc: TRUE
lof: TRUE
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE, include=FALSE}

library(here)
library(lubridate)
library(tidyverse)
library(readr)
library(dplyr)
library(cowplot)
library(corrplot)
library(GGally)
#install.packages("trend")
library(trend)
#install.packages("zoo")
library(zoo)
#install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)

here()
```



## Rationale and Research Question


\newpage

## Dataset Information


\newpage

## Exploration of Raw Data

Exploration of the raw data consisted of understanding the number of files and contents of those files for each Station. Originally, the water quality data and the nutrient data were separate for each station we are studying. We uploaded all of the files with nutrient and water quality data from the four stations we are studying for the years 2002-2024. We obtained a separate file with information about the stations in which the data was collected. 

```{r import of raw data, message=FALSE, warning=FALSE, include=FALSE}
sampling_stations <- read_csv("Data/sampling_stations.csv")

# Importing the data within each folder as one and then combining data from each folder into their own datasets since they have the same columns

# Nutrients
nocecnut <- list.files(path = "./Data/nocecnut", pattern = "\\.csv$", full.names = TRUE)
nocec_nutrients <- nocecnut %>%
  plyr::ldply(read.csv)

noclcnut <- list.files(path = "./Data/noclcnut", pattern = "\\.csv$", full.names = TRUE)
noclc_nutrients <- noclcnut %>%
  plyr::ldply(read.csv)

nocrcnut <- list.files(path = "./Data/nocrcnut", pattern = "\\.csv$", full.names = TRUE)
nocrc_nutrients <- nocrcnut %>%
  plyr::ldply(read.csv)

noczbnut <- list.files(path = "./Data/noczbnut", pattern = "\\.csv$", full.names = TRUE)
noczb_nutrients <- noczbnut %>%
  plyr::ldply(read.csv)

# Water Quality 
nocecwq <- list.files(path = "./Data/nocecwq", pattern = "\\.csv$", full.names = TRUE)
nocec_water_quality <- nocecwq %>%
  plyr::ldply(read.csv)

noclcwq <- list.files(path = "./Data/noclcwq", pattern = "\\.csv$", full.names = TRUE)
noclc_water_quality <- noclcwq %>%
  plyr::ldply(read.csv)

nocrcwq <- list.files(path = "./Data/nocrcwq", pattern = "\\.csv$", full.names = TRUE)
nocrc_water_quality <- nocrcwq %>%
  plyr::ldply(read.csv)

noczbwq <- list.files(path = "./Data/noczbwq", pattern = "\\.csv$", full.names = TRUE)
noczb_water_quality <- noczbwq %>%
  plyr::ldply(read.csv)
```


\newpage

## Data Wrangling

After the conclusion of exploring the raw data, we began to wrangle the data by choosing the variables we wished to keep for our study. For the nutrient data sets we kept:
* Station Code
* Date and Time
* Orthophosphate (PO4F) (mg/L) 
* Ammonnium Fluoride (NH4F) (mg/L) 
* Nitryl Fluoride (NO2F) (mg/L) 
* Fluorine Nitrate (NO3F) (mg/L) 
* Nitrate + Nitrate (NO23F) (mg/L)
* Chlorophyll (ug/L)
For the water quality data sets we kept the variables:  
* Station Code
* Date and Time
* Temperature (degrees Celsius)
* Salinity (ppt)
* Dissolved Oxygen (%)
* Dissolved Oxygen in milligrams per Liter (mg/L)
* Depth (m)
* pH (standard units)
* Turbidity (NTU)

 
```{r first cleaning/choosing the columns, message=FALSE, warning=FALSE, include=FALSE}
# Nutrients
nocec_nutrients_cleaned <- nocec_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F, NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

noclc_nutrients_cleaned <- noclc_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F, NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

nocrc_nutrients_cleaned <- nocrc_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F,NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

noczb_nutrients_cleaned <- noczb_nutrients %>%
  select(StationCode, DateTimeStamp, PO4F, NH4F, 
         NO2F, NO3F, NO23F, CHLA_N)

# Water Quality 
nocec_water_quality_cleaned <- nocec_water_quality %>%
  select(StationCode, DateTimeStamp, Temp, Sal, DO_Pct, 
         DO_mgl, Depth, pH, Turb)

noclc_water_quality_cleaned <- noclc_water_quality %>%
  select(StationCode, DateTimeStamp, Temp, Sal,  DO_Pct, 
         DO_mgl, Depth, pH, Turb)

nocrc_water_quality_cleaned <- nocrc_water_quality %>%
  select(StationCode, DateTimeStamp, Temp, Sal,  DO_Pct, 
         DO_mgl, Depth, pH, Turb)

noczb_water_quality_cleaned <- noczb_water_quality %>% 
  select(StationCode, DateTimeStamp, Temp, Sal,  DO_Pct, 
         DO_mgl, Depth, pH, Turb)
```

```{r combining the datasets, echo=FALSE, message=FALSE, warning=FALSE}
# Nutrients
nutrients <- bind_rows(nocec_nutrients_cleaned,
                       noclc_nutrients_cleaned,
                       nocrc_nutrients_cleaned,
                       noczb_nutrients_cleaned)
str(nutrients)

# Water Quality
water_quality <- bind_rows(nocec_water_quality_cleaned,
                           noclc_water_quality_cleaned,
                           nocrc_water_quality_cleaned,
                           noczb_water_quality_cleaned)
str(water_quality)
```

```{r creating a Date column, message=FALSE, warning=FALSE, include=FALSE}
# Nutrients
nutrients <- nutrients %>%
  mutate(DateTimeStamp = mdy_hm(DateTimeStamp))
  
nutrients <- nutrients %>% 
  mutate(Date = as.Date(DateTimeStamp))

# Water Quality
water_quality <- water_quality %>%
  mutate(DateTimeStamp = mdy_hm(DateTimeStamp))

water_quality <- water_quality %>%
  mutate(Date = as.Date(DateTimeStamp))
```

```{r clean Sampling Station dataset, echo=FALSE, message=FALSE, warning=FALSE}
# Sampling Stations
# rename the  columns bc they have spaces
sampling_stations <- sampling_stations %>%
  rename(NERR_SiteID = `NERR Site ID`,
    StationCode = `Station Code`,
    StationName = `Station Name`,
    LatLong = `Lat Long`,
    Reserve_Name = `Reserve Name`)

#Choosing the columns we want
sampling_stations_cleaned <- sampling_stations %>%
  select(NERR_SiteID, StationCode, StationName, LatLong, 
         Latitude, Longitude, Status, State, Reserve_Name)

sampling_stations_cleaned <- sampling_stations_cleaned %>%
  filter(StationCode %in% c("nocecnut", "nocecwq", "noclcnut", "noclcwq", 
                            "nocrcnut", "nocrcwq", "noczbnut", "noczbwq"))
str(sampling_stations_cleaned)
```

```{r clean up formatting of StationCodes}
# Clean up spaces and formatting differences
nutrients$StationCode <- trimws(nutrients$StationCode)
water_quality$StationCode <- trimws(water_quality$StationCode)
sampling_stations_cleaned$StationCode <- trimws(sampling_stations_cleaned$StationCode)

str(nutrients$StationCode)
str(water_quality$StationCode)
str(sampling_stations_cleaned$StationCode)
```

```{r Averaging the data per day, message=FALSE, warning=FALSE, include=FALSE}
# Nutrient - daily avg of all variables
nutrients_dailyavg <- nutrients %>%
  group_by(Date, StationCode) %>%
  summarise(PO4F = mean(PO4F, na.rm = TRUE),
            NH4F = mean(NH4F, na.rm = TRUE),
            NO2F = mean(NO2F, na.rm = TRUE),
            NO3F = mean(NO3F, na.rm = TRUE),
            NO23F = mean(NO23F, na.rm = TRUE),
            CHLA_N = mean(CHLA_N, na.rm = TRUE))

# Nutrient - monthly avg of all variables 
nutrients_monthlyavg <- nutrients %>%
  mutate(Year = year(Date),
         Month = month(Date)) %>%
  group_by(Year, Month, StationCode) %>%
  summarise(PO4F = mean(PO4F, na.rm = TRUE),
            NH4F = mean(NH4F, na.rm = TRUE),
            NO2F = mean(NO2F, na.rm = TRUE),
            NO3F = mean(NO3F, na.rm = TRUE),
            NO23F = mean(NO23F, na.rm = TRUE),
            CHLA_N = mean(CHLA_N, na.rm = TRUE))
  
# Water Quality - avg per day of all variables
water_quality_dailyavg <- water_quality %>%
  group_by(Date, StationCode) %>%
  summarise(Temp = mean(Temp, na.rm = TRUE),
            Sal = mean(Sal, na.rm = TRUE),
            DO_Pct = mean(DO_Pct, na.rm = TRUE),
            DO_mgl = mean(DO_mgl, na.rm = TRUE),
            Depth = mean(Depth, na.rm = TRUE),
            pH = mean(pH, na.rm = TRUE),
            Turb = mean(Turb, na.rm = TRUE))

# Water Quality - monthly avg of all variables 
water_quality_monthlyavg <- water_quality %>%
  mutate(Year = year(Date),
         Month = month(Date)) %>%
  group_by(Year, Month, StationCode) %>%
  summarise(Temp = mean(Temp, na.rm = TRUE),
            Sal = mean(Sal, na.rm = TRUE),
            DO_Pct = mean(DO_Pct, na.rm = TRUE),
            DO_mgl = mean(DO_mgl, na.rm = TRUE),
            Depth = mean(Depth, na.rm = TRUE),
            pH = mean(pH, na.rm = TRUE),
            Turb = mean(Turb, na.rm = TRUE))
```

After cleaning, preparing, and combining all of the files into one data set for nutrients and one data set for water quality, we added the sampling station data to each. This allowed us to create two large data sets which we will use for our study.

```{r add station names to nutrient and water quality data, message=FALSE, warning=FALSE, include=FALSE}
#joining station names to nutrients
nutrients_stations_dailyavg <- 
  left_join(nutrients_dailyavg, sampling_stations_cleaned, by = "StationCode")

nutrients_stations_monthlyavg <- 
  left_join(nutrients_monthlyavg, sampling_stations_cleaned, by = "StationCode")

#joining station names to water_quality
water_quality_stations <- 
  left_join(water_quality_dailyavg, sampling_stations_cleaned, by = "StationCode")

water_quality_stations <- 
  left_join(water_quality_monthlyavg, sampling_stations_cleaned, by = "StationCode")
```

```{r}
#creating a proper date column for time series analysis
water_quality_stations <- water_quality_stations %>% 
  mutate(Date = ym(paste(Year, Month)))
nutrients_stations_monthlyavg <- nutrients_stations_monthlyavg %>% 
  mutate(Date = ym(paste(Year,Month)))
```

```{r}
#merging the two datasets to make one monstrous master dataset
merged_test <- full_join(water_quality_stations, nutrients_stations_monthlyavg,
                    by = c("StationName", "Date"),
                    suffix = c(".wq", ".nutr"))

#removing duplicated columns
stations_wq_nutr_cleaned <- merged_test %>%
  mutate(
    StationCode = coalesce(StationCode.wq, StationCode.nutr),
    Latitude = coalesce(Latitude.wq, Latitude.nutr),
    Longitude = coalesce(Longitude.wq, Longitude.nutr),
    LatLong = coalesce(LatLong.wq, LatLong.nutr),
    Year = coalesce(Year.wq, Year.nutr),
    Month = coalesce(Month.wq, Month.nutr),
    NERR_SiteID = coalesce(NERR_SiteID.wq, NERR_SiteID.nutr),
    Status = coalesce(Status.wq, Status.nutr),
    State = coalesce(State.wq, State.nutr),
    Reserve_Name = coalesce(Reserve_Name.wq, Reserve_Name.nutr),
  ) %>%
  select(-StationCode.wq, -StationCode.nutr,
         -Latitude.wq, -Latitude.nutr,
         -Longitude.wq, -Longitude.nutr,
         -Year.wq, -Year.nutr,
         -Month.wq, -Month.nutr,
         -NERR_SiteID.wq, -NERR_SiteID.nutr,
         -Status.wq, -Status.nutr,
         -State.wq, -State.nutr,
         -Reserve_Name.wq, -Reserve_Name.nutr,
         -LatLong.wq, -LatLong.nutr) %>% 
  arrange(StationName, Date)
```


Now we want to save this final cleaned data set and make sure it's pushed to our repository in Github!
```{r}
write.csv(water_quality_stations, row.names = FALSE, file =
            "./Data/water_quality_stations_monthlyavg_cleaned")
write.csv(nutrients_stations_monthlyavg, row.names = FALSE, file =
            "./Data/nutrients_stations_monthlyavg_cleaned")
write.csv(stations_wq_nutr_cleaned, row.names = FALSE, file =
            "./Data/stations_wq_nutr_cleaned")
```


\newpage

## Exploration of Processed Data

```{r}
#Setting our theme, a lovely light toothpaste seafoam green to match a eutrophication aesthetic?
our_theme <- theme_classic(base_size = 14, base_family = "serif") +
  theme(plot.title = element_text(color = '#4F4F4F'),
        axis.ticks = element_line(color = '#4F4F4F'),
        axis.line = element_line(color = '#ffffff'),
        axis.title.x = element_text(color = '#4F4F4F'),
        axis.title.y = element_text(color = '#4F4F4F'),
        axis.text = element_text(color = '#4F4F4F'),
        legend.text = element_text(color = '#4F4F4F'),
        legend.title = element_text(color = '#4F4F4F'),
        legend.background = element_rect(fill = '#ffffff'),
        plot.background = element_rect(fill = '#ffffff'),
        panel.background = element_rect(fill = '#cadcd3'),
        panel.grid.major = element_line(color = "#ffffff"), 
        panel.grid.minor = element_line(color = "#ffffff"),
        rect = element_rect(color = "#ffffff"))
```

Building Correlation Matrix
```{r}
num_data <- stations_wq_nutr_cleaned[, c("Temp", "DO_mgl", "Depth",
                                         "pH", "PO4F", "NH4F", "NO3F",
                                         "CHLA_N", "Month")]
cor_matrix <- cor(num_data, use = "pairwise.complete.obs", method = "pearson")

```

ggpairs
```{r, warning=FALSE}
ggpairs(num_data)
```

corrplot, something more readable than ggpairs
```{r}
corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```


We're most interested in Dissolved Oxygen across the different monitoring stations. Is there consistent data for us to analyze? Could we identify trends over time?
```{r}
#looking at dissolved oxygen over time for each station
stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = Date, y = DO_mgl, color = StationName)) +
  geom_line() +
  xlab("Date") +
  ylab("Dissolved Oxygen (mgl)") +
  labs(color = "Station Name") +
  our_theme
```

Although plotting DO is helpful for seeing consistent patterns and concentrations across all four stations, let's look at them separately to see gaps and data coverage.
```{r}
#Using facets to plot DO separately for each monitoring station
stations_wq_nutr_cleaned %>%
  ggplot(mapping = aes(x=Date, y=DO_mgl)) +
  geom_line(aes(color = StationName),) +
  xlab("Date") +
  ylab("Dissolved Oxygen (mgl)") +
  theme(axis.text.x = (element_text(size=7))) +
  facet_wrap(vars(StationName), nrow = 4) +
  our_theme
```
East Cribbing's data seems to end by 2022. This is good to know for our analysis.

It seems that there's variation in dissolved oxygen across the year. What do the monthly spreads look like?
```{r}
#looking at monthly averages
stations_wq_nutr_cleaned %>%
  ggplot(aes(x = factor(Month,levels=1:12,labels=month.abb), y = DO_mgl, color = StationName)) +
  geom_boxplot() +
  xlab("Date") +
  ylab("Dissolved Oxygen (mgl)") +
  labs(color = "Station Name") +
  our_theme
```

What's the relationship between temperature and dissolved oxygen?
```{r}
Temp_x_limits <- c(0, 35)

Temp_DO_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = Temp, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Temperature (Celsius)") +
  ylab("Dissolved Oxygen (mg/L)") +
  scale_x_continuous(limits = Temp_x_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

Temp_DO_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = Temp, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Temperature (Celsius)") +
  ylab(NULL) +
  scale_x_continuous(limits = Temp_x_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  Temp_DO_combined,
  Temp_DO_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```


What's the relationship between nitrate and dissolved oxygen?
```{r}
NO3F_DO_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = NO3F, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Nitrate") +
  ylab("Dissolved Oxygen (mg/L)") +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

NO3F_DO_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = NO3F, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Nitrate") +
  ylab(NULL) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  NO3F_DO_combined,
  NO3F_DO_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

What's the relationship between phosphate and dissolved oxygen?
```{r}
PO4F_x_limits <- c(-0.001, 0.15)

PO4F_DO_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = PO4F, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm, formula = y ~ x, se = FALSE, aes(color = NULL)) +
  xlab("PO4F") +
  ylab("Dissolved Oxygen (mg/L)") +
  scale_x_continuous(limits = PO4F_x_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

PO4F_DO_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = PO4F, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm, formula = y ~ x, se = FALSE, aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("PO4F") +
  ylab(NULL) +
  scale_x_continuous(limits = PO4F_x_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  PO4F_DO_combined,
  PO4F_DO_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

Ammonium?
```{r}
NH4F_x_limits <- c(-0.01, 0.5)

NH4F_DO_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = NH4F, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Ammonium") +
  ylab("Dissolved Oxygen (mg/L)") +
  scale_x_continuous(limits = NH4F_x_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

NH4F_DO_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = NH4F, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Ammonium") +
  ylab(NULL) +
  scale_x_continuous(limits = NH4F_x_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  NH4F_DO_combined,
  NH4F_DO_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```


Depth
What's the relationship between temperature and dissolved oxygen?
```{r}
Depth_DO_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = Depth, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Depth") +
  ylab("Dissolved Oxygen (mg/L)") +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

Depth_DO_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = Depth, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Depth") +
  ylab(NULL) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  Depth_DO_combined,
  Depth_DO_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

We are also interested in Chlorophyll, as it can indicate algal blooms.

```{r}
#looking at chlorophyll over time for each station
stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = Date, y = CHLA_N, color = StationName)) +
  geom_line() +
  xlab("Date") +
  ylab("CHLA_N") +
  labs(color = "Station Name") +
  our_theme
```

Although plotting Chlorophyll is helpful for seeing consistent patterns and concentrations across all four stations, let's look at them separately to see gaps and data coverage.
```{r}
#Using facets to plot chlorophyll separately for each monitoring station
stations_wq_nutr_cleaned %>%
  ggplot(mapping = aes(x=Date, y=CHLA_N)) +
  geom_line(aes(color = StationName),) +
  xlab("Date") +
  ylab("CHLA_N") +
  theme(axis.text.x = (element_text(size=7))) +
  facet_wrap(vars(StationName), nrow = 4) +
  scale_y_continuous(limits = c(0,75)) +
  our_theme
```
Loosin Creek and Research Creek seem to have less fluctuation in their Chlorophyll than East Cribbbing and Zeke's Basin. Could it do with differences in depth or other variables?

It seems that there's variation in chlorophyll across the year. What do the monthly spreads look like?
```{r}
#looking at monthly averages
stations_wq_nutr_cleaned %>%
  ggplot(aes(x = factor(Month,levels=1:12,labels=month.abb), y = CHLA_N, color = StationName)) +
  geom_boxplot() +
  xlab("Date") +
  ylab("CHLA_N") +
  labs(color = "Station Name") +
  scale_y_continuous(limits = c(0,75)) +
  our_theme
```

What's the relationship between temperature and chlorophpyll?
```{r}
#Looking at temperature vs dissolved oxygen with trend line
Temp_x_limits <- c(0, 35)
CHLA_N_y_limits <- c(0,75)

Temp_CHLA_N_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = Temp, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Temperature (Celsius)") +
  ylab("CHLA_N") +
  scale_x_continuous(limits = Temp_x_limits) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

Temp_CHLA_N_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = Temp, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Temperature (Celsius)") +
  ylab(NULL) +
  scale_x_continuous(limits = Temp_x_limits) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  Temp_CHLA_N_combined,
  Temp_CHLA_N_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

What's the relationship between nitrate and dissolved oxygen?
```{r}
#Looking at nitrate vs chlorophyll with trend line
CHLA_N_y_limits <- c(0,75)

NO3F_CHLA_N_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = NO3F, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Nitrate") +
  ylab("CHLA_N") +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

NO3F_CHLA_N_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = NO3F, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Nitrate") +
  ylab(NULL) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  NO3F_CHLA_N_combined,
  NO3F_CHLA_N_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

What's the relationship between phosphate and dissolved oxygen?
```{r}
#looking at phosphate vs chlorophyll with trend line
PO4F_x_limits <- c(-0.001, 0.15)
CHLA_N_y_limits <- c(0,75)

PO4F_CHLA_N_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = PO4F, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Phosphate") +
  ylab("CHLA_N") +
  scale_x_continuous(limits = PO4F_x_limits) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

PO4F_CHLA_N_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = PO4F, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Phosphate") +
  ylab(NULL) +
  scale_x_continuous(limits = PO4F_x_limits) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  PO4F_CHLA_N_combined,
  PO4F_CHLA_N_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

Ammonium?
```{r}
#Looking at ammonium vs chlorophyll with trend line
NH4F_x_limits <- c(-0.01, 0.5)
CHLA_N_y_limits <- c(0,75)

NH4F_CHLA_N_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = NH4F, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Ammonium") +
  ylab("CHLA_N") +
  scale_x_continuous(limits = NH4F_x_limits) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

NH4F_CHLA_N_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = NH4F, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Ammonium") +
  ylab(NULL) +
  scale_x_continuous(limits = NH4F_x_limits) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  NH4F_CHLA_N_combined,
  NH4F_CHLA_N_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

Depth
```{r}
#Looking at depth vs chlorophyll with trend line
CHLA_N_y_limits <- c(0,75)

Depth_CHLA_N_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = Depth, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("Depth") +
  ylab("CHLA_N") +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

Depth_CHLA_N_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = Depth, y = CHLA_N, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("Depth") +
  ylab(NULL) +
  scale_y_continuous(limits = CHLA_N_y_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  Depth_CHLA_N_combined,
  Depth_CHLA_N_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```

Now, what does the relationship between dissolved oxygen and chlorophyll look like?
```{r}
#Looking at dissolved oxygen vs chlorophyll with trend line
CHLA_N_x_limits <- c(0,75)

DO_CHLA_N_combined <- stations_wq_nutr_cleaned %>% 
  ggplot(aes(x = CHLA_N, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.8, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x, 
              se = FALSE, aes(color = NULL)) +
  xlab("CHLA_N") +
  ylab("Dissolved Oxygen (mg/L)") +
  scale_x_continuous(limits = CHLA_N_x_limits) +
  our_theme +
  theme(legend.position = "none")      # legend on right for combined figure

DO_CHLA_N_faceted <- stations_wq_nutr_cleaned %>%
  ggplot(aes(x = CHLA_N, y = DO_mgl, color = StationName)) +
  geom_point(size = 1.4, alpha = 0.8) +
  geom_smooth(method = MASS::rlm,
              formula = y ~ x,
              se = FALSE,
              aes(color = NULL)) +
  facet_wrap(vars(StationName), nrow = 4) +
  xlab("CHLA_N") +
  ylab(NULL) +
  scale_x_continuous(limits = CHLA_N_x_limits) +
  our_theme +
  theme(legend.position = "none")

plot_grid(
  DO_CHLA_N_combined,
  DO_CHLA_N_faceted,
  ncol = 2,          # side-by-side layout
  rel_widths = c(1, 1)  # adjust width proportions if needed
)
```


\newpage

## Analysis
We want to do two main things: 1) Identify if there is a trend in decreasing dissolved oxygen and increasing chlorophyll across time, and 2) identify the variables most strongly correlated with possible changes.

First, we must turn dissolved oxygen into a time series across each station.

Let's start with East Cribbing.
```{r}
#selecting just East Cribbing's DO data and filtering out years with no data
EastCribbing_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "East Cribbing") %>% 
  select(Date, DO_mgl, DO_Pct, Year, Month) %>% 
  filter(Year < 2022)

#Finding gaps by making new data frame
EastCribbing_months <- as.data.frame(seq.Date(from=as.Date("2002/01/01"),
                                 to=as.Date("2021/12/01"),
                                 by="month"))
colnames(EastCribbing_months) <- "Date"

#Left joining to point out gaps
EastCribbing_timeseries_ready <- left_join(EastCribbing_months, EastCribbing_monthly, by = "Date")

#Using linear interpolation to fill in DO
EastCribbing_timeseries_ready$DO_mgl = 
  zoo::na.approx(EastCribbing_timeseries_ready$DO_mgl, rule = 2)
EastCribbing_timeseries_ready$DO_Pct = 
  zoo::na.approx(EastCribbing_timeseries_ready$DO_Pct, rule = 2)

#Creating time series for East Cribbing's DO
EastCribbingDO.monthly.ts <- ts(EastCribbing_timeseries_ready$DO_mgl, frequency = 12, start = c(2002,1))

#Decomposing
EastCribbingDO.monthly.decomposed <- stl(EastCribbingDO.monthly.ts, s.window = "periodic")

#plotting
plot(EastCribbingDO.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
EastCribbingDO.monthly.trend <- 
  Kendall::SeasonalMannKendall(EastCribbingDO.monthly.ts)

summary(EastCribbingDO.monthly.trend)

#subtracting seasonal component from our ts object
EastCribbingDO.monthly.nonseas <- (EastCribbingDO.monthly.ts - EastCribbingDO.monthly.decomposed$time.series[,1])

#making sure it worked
plot(EastCribbingDO.monthly.nonseas)
plot(EastCribbingDO.monthly.ts)

#16: Running Mann Kendall on non-seasonal Ozone monthly series
EastCribbingDO.monthly.nonseas.trend <- trend::mk.test(EastCribbingDO.monthly.nonseas)
# Inspect results
EastCribbingDO.monthly.nonseas.trend
```

now lets do Loosin Creek
```{r}
#selecting just Loosin Creek's DO data and filtering out years with no data
LoosinCreek_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Loosin Creek") %>% 
  select(Date, DO_mgl, DO_Pct, Year, Month)

#Finding gaps by making new data frame
LoosinCreek_months <- as.data.frame(seq.Date(from=as.Date("2002/02/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(LoosinCreek_months) <- "Date"

#Left joining to point out gaps
LoosinCreek_timeseries_ready <- left_join(LoosinCreek_months, LoosinCreek_monthly, by = "Date")

#Using linear interpolation to fill in DO
LoosinCreek_timeseries_ready$DO_mgl = 
  zoo::na.approx(LoosinCreek_timeseries_ready$DO_mgl)
LoosinCreek_timeseries_ready$DO_Pct = 
  zoo::na.approx(LoosinCreek_timeseries_ready$DO_Pct)

#Creating time series for Loosin Creek's DO
LoosinCreekDO.monthly.ts <- ts(LoosinCreek_timeseries_ready$DO_mgl, frequency = 12, start = c(2002,2))

#Decomposing
LoosinCreekDO.monthly.decomposed <- stl(LoosinCreekDO.monthly.ts, s.window = "periodic")

#plotting
plot(LoosinCreekDO.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
LoosinCreekDO.monthly.trend <- 
  Kendall::SeasonalMannKendall(LoosinCreekDO.monthly.ts)

summary(LoosinCreekDO.monthly.trend)

#subtracting seasonal component from our ts object
LoosinCreekDO.monthly.nonseas <- (LoosinCreekDO.monthly.ts - LoosinCreekDO.monthly.decomposed$time.series[,1])

#making sure it worked
plot(LoosinCreekDO.monthly.nonseas)
plot(LoosinCreekDO.monthly.ts)

#Running Mann Kendall on non-seasonal monthly series
LoosinCreekDO.monthly.nonseas.trend <- trend::mk.test(LoosinCreekDO.monthly.nonseas)
# Inspect results
LoosinCreekDO.monthly.nonseas.trend
```

Now Research Creek
```{r}
#selecting just Research Creek's DO data and filtering out years with no data
ResearchCreek_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Research Creek") %>% 
  select(Date, DO_mgl, DO_Pct, Year, Month)

#Finding gaps by making new data frame
ResearchCreek_months <- as.data.frame(seq.Date(from=as.Date("2002/01/01"),
                                 to=as.Date("2025/03/01"),
                                 by="month"))
colnames(ResearchCreek_months) <- "Date"

#Left joining to point out gaps
ResearchCreek_timeseries_ready <- left_join(ResearchCreek_months, ResearchCreek_monthly, by = "Date")

#Using linear interpolation to fill in DO
ResearchCreek_timeseries_ready$DO_mgl = 
  zoo::na.approx(ResearchCreek_timeseries_ready$DO_mgl)
ResearchCreek_timeseries_ready$DO_Pct = 
  zoo::na.approx(ResearchCreek_timeseries_ready$DO_Pct)

#Creating time series for Research Creek's DO
ResearchCreekDO.monthly.ts <- ts(ResearchCreek_timeseries_ready$DO_mgl, frequency = 12, start = c(2002,1))

#Decomposing
ResearchCreekDO.monthly.decomposed <- stl(ResearchCreekDO.monthly.ts, s.window = "periodic")

#plotting
plot(ResearchCreekDO.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ResearchCreekDO.monthly.trend <- 
  Kendall::SeasonalMannKendall(ResearchCreekDO.monthly.ts)

summary(ResearchCreekDO.monthly.trend)

#subtracting seasonal component from our ts object
ResearchCreekDO.monthly.nonseas <- (ResearchCreekDO.monthly.ts - ResearchCreekDO.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ResearchCreekDO.monthly.nonseas)
plot(ResearchCreekDO.monthly.ts)

#Running Mann Kendall on non-seasonal monthly series
ResearchCreekDO.monthly.nonseas.trend <- trend::mk.test(ResearchCreekDO.monthly.nonseas)
# Inspect results
ResearchCreekDO.monthly.nonseas.trend
```

Lastly, Zeke's Basin
```{r}
#selecting just Research Creek's DO data and filtering out years with no data
ZekesBasin_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Zeke's Basin") %>% 
  select(Date, DO_mgl, DO_Pct, Year, Month)

#Finding gaps by making new data frame
ZekesBasin_months <- as.data.frame(seq.Date(from=as.Date("2002/03/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ZekesBasin_months) <- "Date"

#Left joining to point out gaps
ZekesBasin_timeseries_ready <- left_join(ZekesBasin_months, ZekesBasin_monthly, by = "Date")

#Using linear interpolation to fill in DO
ZekesBasin_timeseries_ready$DO_mgl = 
  zoo::na.approx(ZekesBasin_timeseries_ready$DO_mgl)
ZekesBasin_timeseries_ready$DO_Pct = 
  zoo::na.approx(ZekesBasin_timeseries_ready$DO_Pct)

#Creating time series for Research Creek's DO
ZekesBasinDO.monthly.ts <- ts(ZekesBasin_timeseries_ready$DO_mgl, frequency = 12, start = c(2002,3))

#Decomposing
ZekesBasinDO.monthly.decomposed <- stl(ZekesBasinDO.monthly.ts, s.window = "periodic")

#plotting
plot(ZekesBasinDO.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ZekesBasinDO.monthly.trend <- 
  Kendall::SeasonalMannKendall(ZekesBasinDO.monthly.ts)

summary(ZekesBasinDO.monthly.trend)

#subtracting seasonal component from our ts object
ZekesBasinDO.monthly.nonseas <- (ZekesBasinDO.monthly.ts - ZekesBasinDO.monthly.decomposed$time.series[,1])


#making sure it worked
plot(ZekesBasinDO.monthly.nonseas)
plot(ZekesBasinDO.monthly.ts)

#Running Mann Kendall on non-seasonal monthly series
ZekesBasinDO.monthly.nonseas.trend <- trend::mk.test(ZekesBasinDO.monthly.nonseas)
# Inspect results
ZekesBasinDO.monthly.nonseas.trend
```
All Monitoring Stations have had statistically significant trends in their Dissolved Oxygen.

What about chlorophyll?
Let's start with East Cribbing.
```{r}
#selecting just East Cribbing's CHLA data and filtering out years with no data
EastCribbing_CHLA_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "East Cribbing") %>% 
  select(Date, CHLA_N, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01"))

#Finding gaps by making new data frame
EastCribbing_CHLA_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(EastCribbing_CHLA_months) <- "Date"

#Left joining to point out gaps
EastCribbing_CHLA_timeseries_ready <- left_join(EastCribbing_CHLA_months, EastCribbing_CHLA_monthly, by = "Date")

#Using linear interpolation to fill in DO
EastCribbing_CHLA_timeseries_ready$CHLA_N = 
  zoo::na.approx(EastCribbing_CHLA_timeseries_ready$CHLA_N, rule = 2)

#Creating time series for East Cribbing's DO
EastCribbingCHLA.monthly.ts <- ts(EastCribbing_CHLA_timeseries_ready$CHLA_N, frequency = 12, start = c(2002,4))

#Decomposing
EastCribbingCHLA.monthly.decomposed <- stl(EastCribbingCHLA.monthly.ts, s.window = "periodic")

#plotting
plot(EastCribbingCHLA.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
EastCribbingCHLA.monthly.trend <- 
  Kendall::SeasonalMannKendall(EastCribbingCHLA.monthly.ts)

summary(EastCribbingCHLA.monthly.trend)

#subtracting seasonal component from our ts object
EastCribbingCHLA.monthly.nonseas <- (EastCribbingCHLA.monthly.ts - EastCribbingCHLA.monthly.decomposed$time.series[,1])

#making sure it worked
plot(EastCribbingCHLA.monthly.nonseas)
plot(EastCribbingCHLA.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
EastCribbingCHLA.monthly.nonseas.trend <- trend::mk.test(EastCribbingCHLA.monthly.nonseas)
# Inspect results
EastCribbingCHLA.monthly.nonseas.trend
```

now lets do Loosin Creek
```{r}
#selecting just Loosin Creek's CHLA data and filtering out years with no data
LoosinCreek_CHLA_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Loosin Creek") %>% 
  select(Date, CHLA_N, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01"))

#Finding gaps by making new data frame
LoosinCreek_CHLA_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(LoosinCreek_CHLA_months) <- "Date"

#Left joining to point out gaps
LoosinCreek_CHLA_timeseries_ready <- left_join(LoosinCreek_CHLA_months, LoosinCreek_CHLA_monthly, by = "Date")

#Using linear interpolation to fill in DO
LoosinCreek_CHLA_timeseries_ready$CHLA_N = 
  zoo::na.approx(LoosinCreek_CHLA_timeseries_ready$CHLA_N)

#Creating time series for East Cribbing's DO
LoosinCreekCHLA.monthly.ts <- ts(LoosinCreek_CHLA_timeseries_ready$CHLA_N, frequency = 12, start = c(2002,3))

#Decomposing
LoosinCreekCHLA.monthly.decomposed <- stl(LoosinCreekCHLA.monthly.ts, s.window = "periodic")

#plotting
plot(LoosinCreekCHLA.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
LoosinCreekCHLA.monthly.trend <- 
  Kendall::SeasonalMannKendall(LoosinCreekCHLA.monthly.ts)

summary(LoosinCreekCHLA.monthly.trend)

#subtracting seasonal component from our ts object
LoosinCreekCHLA.monthly.nonseas <- (LoosinCreekCHLA.monthly.ts - LoosinCreekCHLA.monthly.decomposed$time.series[,1])

#making sure it worked
plot(LoosinCreekCHLA.monthly.nonseas)
plot(LoosinCreekCHLA.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
LoosinCreekCHLA.monthly.nonseas.trend <- trend::mk.test(LoosinCreekCHLA.monthly.nonseas)
# Inspect results
LoosinCreekCHLA.monthly.nonseas.trend
```

Now Research Creek
```{r}
#selecting just Research Creek's CHLA data and filtering out years with no data
ResearchCreek_CHLA_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Research Creek") %>% 
  select(Date, CHLA_N, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ResearchCreek_CHLA_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ResearchCreek_CHLA_months) <- "Date"

#Left joining to point out gaps
ResearchCreek_CHLA_timeseries_ready <- left_join(ResearchCreek_CHLA_months, ResearchCreek_CHLA_monthly, by = "Date")

#Using linear interpolation to fill in CHLA
ResearchCreek_CHLA_timeseries_ready$CHLA_N = 
  zoo::na.approx(ResearchCreek_CHLA_timeseries_ready$CHLA_N)

#Creating time series for Research Creek's CHLA
ResearchCreekCHLA.monthly.ts <- ts(ResearchCreek_CHLA_timeseries_ready$CHLA_N, frequency = 12, start = c(2002,3))

#Decomposing
ResearchCreekCHLA.monthly.decomposed <- stl(ResearchCreekCHLA.monthly.ts, s.window = "periodic")

#plotting
plot(ResearchCreekCHLA.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ResearchCreekCHLA.monthly.trend <- 
  Kendall::SeasonalMannKendall(ResearchCreekCHLA.monthly.ts)

summary(ResearchCreekCHLA.monthly.trend)

#subtracting seasonal component from our ts object
ResearchCreekCHLA.monthly.nonseas <- (ResearchCreekCHLA.monthly.ts - ResearchCreekCHLA.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ResearchCreekCHLA.monthly.nonseas)
plot(ResearchCreekCHLA.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ResearchCreekCHLA.monthly.nonseas.trend <- trend::mk.test(ResearchCreekCHLA.monthly.nonseas)
# Inspect results
ResearchCreekCHLA.monthly.nonseas.trend
```

Lastly, Zeke's Basin
```{r}
#selecting just Zeke's Basin's CHLA data and filtering out years with no data
ZekeBasin_CHLA_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Zeke's Basin") %>% 
  select(Date, CHLA_N, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ZekeBasin_CHLA_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ZekeBasin_CHLA_months) <- "Date"

#Left joining to point out gaps
ZekeBasin_CHLA_timeseries_ready <- left_join(ZekeBasin_CHLA_months, ZekeBasin_CHLA_monthly, by = "Date")

#Using linear interpolation to fill in CHLA
ZekeBasin_CHLA_timeseries_ready$CHLA_N = 
  zoo::na.approx(ZekeBasin_CHLA_timeseries_ready$CHLA_N)

#Creating time series for Zeke's Basin's CHLA
ZekeBasinCHLA.monthly.ts <- ts(ZekeBasin_CHLA_timeseries_ready$CHLA_N, frequency = 12, start = c(2002,3))

#Decomposing
ZekeBasinCHLA.monthly.decomposed <- stl(ZekeBasinCHLA.monthly.ts, s.window = "periodic")

#plotting
plot(ZekeBasinCHLA.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ZekeBasinCHLA.monthly.trend <- 
  Kendall::SeasonalMannKendall(ZekeBasinCHLA.monthly.ts)

summary(ZekeBasinCHLA.monthly.trend)

#subtracting seasonal component from our ts object
ZekeBasinCHLA.monthly.nonseas <- (ZekeBasinCHLA.monthly.ts - ZekeBasinCHLA.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ZekeBasinCHLA.monthly.nonseas)
plot(ZekeBasinCHLA.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ZekeBasinCHLA.monthly.nonseas.trend <- trend::mk.test(ZekeBasinCHLA.monthly.nonseas)
# Inspect results
ZekeBasinCHLA.monthly.nonseas.trend
```
All of the stations have statistically significant trends in their DO and Chlorophyll levels. So now we ask, which variables are most responsible for these changes?  

```{r}
#DO
stations_wo_na <- stations_wq_nutr_cleaned[complete.cases(
  stations_wq_nutr_cleaned[, c("DO_mgl","Temp","PO4F","NH4F","NO3F", "CHLA_N")]), ]

DO.AIC <- lm(data = stations_wo_na, DO_mgl~ Temp + PO4F + NH4F + NO3F)
step(DO.AIC)

DO.regression <- lm(data = stations_wq_nutr_cleaned, DO_mgl~ Temp + PO4F
                      + NO3F)
summary(DO.regression)
```
```{r}
CHLA.AIC <- lm(data = stations_wo_na, CHLA_N~ Temp + PO4F + NH4F + NO3F)
step(CHLA.AIC)

CHLA.regression <- lm(data = stations_wq_nutr_cleaned, CHLA_N~ Temp + PO4F
                      + NH4F)
summary(CHLA.regression)
```
From these multiple regressions we find that temperature, phosphorus, and nitrate are the best variables for predicting DO and temperature, phosphorus, and ammonium are the best variables for predicting Chlorophyll. Both of these relationships are significant, although the R-squared for the Chlorophyll regression is 0.1661 which indicates a weaker correlation. 

Temperature is the most impactful variable in both of the models, with PO4 being the least impactful. 

From here, we perform time series analyses on these variables to determine if their trends are in line with what we expect. 

Here are the variables for East Cribbing. 
```{r}
#PO4 - East Cribbing
EastCribbing_PO4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "East Cribbing") %>% 
  select(Date, PO4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
EastCribbing_PO4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(EastCribbing_PO4_months) <- "Date"

#Left joining to point out gaps
EastCribbing_PO4_timeseries_ready <- left_join(EastCribbing_PO4_months, 
                                               EastCribbing_PO4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in PO4
EastCribbing_PO4_timeseries_ready$PO4 = 
  zoo::na.approx(EastCribbing_PO4_timeseries_ready$PO4)

#Creating time series for East Cribbing's PO4
EastCribbing.PO4.monthly.ts <- ts(EastCribbing_PO4_timeseries_ready$PO4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
EastCribbing.PO4.monthly.decomposed <- stl(EastCribbing.PO4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(EastCribbing.PO4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
EastCribbing.PO4.monthly.trend <- 
  Kendall::SeasonalMannKendall(EastCribbing.PO4.monthly.ts)

summary(EastCribbing.PO4.monthly.trend)

#subtracting seasonal component from our ts object
EastCribbing.PO4.monthly.nonseas <- (EastCribbing.PO4.monthly.ts - 
                                       EastCribbing.PO4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(EastCribbing.PO4.monthly.nonseas)
plot(EastCribbing.PO4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
EastCribbing.PO4.monthly.nonseas.trend <- trend::mk.test(EastCribbing.PO4.monthly.nonseas)
# Inspect results
EastCribbing.PO4.monthly.nonseas.trend
```
Increasing trend of PO4, significant. 
```{r}
#NH4 - East Cribbing
EastCribbing_NH4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "East Cribbing") %>% 
  select(Date, NH4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
EastCribbing_NH4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(EastCribbing_NH4_months) <- "Date"

#Left joining to point out gaps
EastCribbing_NH4_timeseries_ready <- left_join(EastCribbing_NH4_months, 
                                               EastCribbing_NH4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in CHLA
EastCribbing_NH4_timeseries_ready$NH4 = 
  zoo::na.approx(EastCribbing_NH4_timeseries_ready$NH4)

#Creating time series for Zeke's Basin's CHLA
EastCribbing.NH4.monthly.ts <- ts(EastCribbing_NH4_timeseries_ready$NH4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
EastCribbing.NH4.monthly.decomposed <- stl(EastCribbing.NH4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(EastCribbing.NH4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
EastCribbing.NH4.monthly.trend <- 
  Kendall::SeasonalMannKendall(EastCribbing.NH4.monthly.ts)

summary(EastCribbing.NH4.monthly.trend)

#subtracting seasonal component from our ts object
EastCribbing.NH4.monthly.nonseas <- (EastCribbing.NH4.monthly.ts - 
                                       EastCribbing.NH4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(EastCribbing.NH4.monthly.nonseas)
plot(EastCribbing.NH4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
EastCribbing.NH4.monthly.nonseas.trend <- trend::mk.test(EastCribbing.NH4.monthly.nonseas)
# Inspect results
EastCribbing.NH4.monthly.nonseas.trend
```
Decreasing trend of NH4, significant. 
```{r}
#NO3 - East Cribbing
EastCribbing_NO3_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "East Cribbing") %>% 
  select(Date, NO3F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
EastCribbing_NO3_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(EastCribbing_NO3_months) <- "Date"

#Left joining to point out gaps
EastCribbing_NO3_timeseries_ready <- left_join(EastCribbing_NO3_months, 
                                               EastCribbing_NO3_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in CHLA
EastCribbing_NO3_timeseries_ready$NO3 = 
  zoo::na.approx(EastCribbing_NO3_timeseries_ready$NO3)

#Creating time series for Zeke's Basin's CHLA
EastCribbing.NO3.monthly.ts <- ts(EastCribbing_NO3_timeseries_ready$NO3, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
EastCribbing.NO3.monthly.decomposed <- stl(EastCribbing.NO3.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(EastCribbing.NO3.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
EastCribbing.NO3.monthly.trend <- 
  Kendall::SeasonalMannKendall(EastCribbing.NO3.monthly.ts)

summary(EastCribbing.NO3.monthly.trend)

#subtracting seasonal component from our ts object
EastCribbing.NO3.monthly.nonseas <- (EastCribbing.NO3.monthly.ts - 
                                       EastCribbing.NO3.monthly.decomposed$time.series[,1])

#making sure it worked
plot(EastCribbing.NO3.monthly.nonseas)
plot(EastCribbing.NO3.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
EastCribbing.NO3.monthly.nonseas.trend <- trend::mk.test(EastCribbing.NO3.monthly.nonseas)
# Inspect results
EastCribbing.NO3.monthly.nonseas.trend
```
Insignificant trend. 
```{r}
#Temp - East Cribbing
EastCribbing_Temp_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "East Cribbing") %>% 
  select(Date, Temp, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
EastCribbing_Temp_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2021/11/01"),
                                 by="month"))
colnames(EastCribbing_Temp_months) <- "Date"

#Left joining to point out gaps
EastCribbing_Temp_timeseries_ready <- left_join(EastCribbing_Temp_months, 
                                               EastCribbing_Temp_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in Temp
EastCribbing_Temp_timeseries_ready$Temp = 
  zoo::na.approx(EastCribbing_Temp_timeseries_ready$Temp)

#Creating time series for East Cribbing's Temp
EastCribbing.Temp.monthly.ts <- ts(EastCribbing_Temp_timeseries_ready$Temp, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
EastCribbing.Temp.monthly.decomposed <- stl(EastCribbing.Temp.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(EastCribbing.Temp.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
EastCribbing.Temp.monthly.trend <- 
  Kendall::SeasonalMannKendall(EastCribbing.Temp.monthly.ts)

summary(EastCribbing.Temp.monthly.trend)

#subtracting seasonal component from our ts object
EastCribbing.Temp.monthly.nonseas <- (EastCribbing.Temp.monthly.ts - 
                                       EastCribbing.Temp.monthly.decomposed$time.series[,1])

#making sure it worked
plot(EastCribbing.Temp.monthly.nonseas)
plot(EastCribbing.Temp.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
EastCribbing.Temp.monthly.nonseas.trend <- trend::mk.test(EastCribbing.Temp.monthly.nonseas)
# Inspect results
EastCribbing.Temp.monthly.nonseas.trend
```
Insignificant trend. 

For East Cribbing, PO4 had a significant increasing trend and NH4 has a significant decreasing trend. 

Now let's do Loosin Creek. 
```{r}
#PO4 - Loosin Creek
LoosinCreek_PO4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Loosin Creek") %>% 
  select(Date, PO4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
LoosinCreek_PO4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(LoosinCreek_PO4_months) <- "Date"

#Left joining to point out gaps
LoosinCreek_PO4_timeseries_ready <- left_join(LoosinCreek_PO4_months, 
                                               LoosinCreek_PO4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in PO4
LoosinCreek_PO4_timeseries_ready$PO4 = 
  zoo::na.approx(LoosinCreek_PO4_timeseries_ready$PO4)

#Creating time series for Loosin Creek's PO4
LoosinCreek.PO4.monthly.ts <- ts(LoosinCreek_PO4_timeseries_ready$PO4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
LoosinCreek.PO4.monthly.decomposed <- stl(LoosinCreek.PO4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(LoosinCreek.PO4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
LoosinCreek.PO4.monthly.trend <- 
  Kendall::SeasonalMannKendall(LoosinCreek.PO4.monthly.ts)

summary(LoosinCreek.PO4.monthly.trend)

#subtracting seasonal component from our ts object
LoosinCreek.PO4.monthly.nonseas <- (LoosinCreek.PO4.monthly.ts - 
                                       LoosinCreek.PO4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(LoosinCreek.PO4.monthly.nonseas)
plot(LoosinCreek.PO4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
LoosinCreek.PO4.monthly.nonseas.trend <- trend::mk.test(LoosinCreek.PO4.monthly.nonseas)
# Inspect results
LoosinCreek.PO4.monthly.nonseas.trend
```
Non seasonal trend is significantly decreasing. 
```{r}
#NH4 - Loosin Creek
LoosinCreek_NH4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Loosin Creek") %>% 
  select(Date, NH4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
LoosinCreek_NH4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(LoosinCreek_NH4_months) <- "Date"

#Left joining to point out gaps
LoosinCreek_NH4_timeseries_ready <- left_join(LoosinCreek_NH4_months, 
                                               LoosinCreek_NH4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in NH4
LoosinCreek_NH4_timeseries_ready$NH4 = 
  zoo::na.approx(LoosinCreek_NH4_timeseries_ready$NH4)

#Creating time series for Loosin Creek's NH4
LoosinCreek.NH4.monthly.ts <- ts(LoosinCreek_NH4_timeseries_ready$NH4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
LoosinCreek.NH4.monthly.decomposed <- stl(LoosinCreek.NH4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(LoosinCreek.NH4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
LoosinCreek.NH4.monthly.trend <- 
  Kendall::SeasonalMannKendall(LoosinCreek.NH4.monthly.ts)

summary(LoosinCreek.NH4.monthly.trend)

#subtracting seasonal component from our ts object
LoosinCreek.NH4.monthly.nonseas <- (LoosinCreek.NH4.monthly.ts - 
                                       LoosinCreek.NH4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(LoosinCreek.NH4.monthly.nonseas)
plot(LoosinCreek.NH4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
LoosinCreek.NH4.monthly.nonseas.trend <- trend::mk.test(LoosinCreek.NH4.monthly.nonseas)
# Inspect results
LoosinCreek.NH4.monthly.nonseas.trend
```
Decreasing trend, significant. 
```{r}
#NO3 - Loosin Creek
LoosinCreek_NO3_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Loosin Creek") %>% 
  select(Date, NO3F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
LoosinCreek_NO3_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(LoosinCreek_NO3_months) <- "Date"

#Left joining to point out gaps
LoosinCreek_NO3_timeseries_ready <- left_join(LoosinCreek_NO3_months, 
                                               LoosinCreek_NO3_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in NO3
LoosinCreek_NO3_timeseries_ready$NO3 = 
  zoo::na.approx(LoosinCreek_NO3_timeseries_ready$NO3)

#Creating time series for Loosin Creek's NO3
LoosinCreek.NO3.monthly.ts <- ts(LoosinCreek_NO3_timeseries_ready$NO3, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
LoosinCreek.NO3.monthly.decomposed <- stl(LoosinCreek.NO3.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(LoosinCreek.NO3.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
LoosinCreek.NO3.monthly.trend <- 
  Kendall::SeasonalMannKendall(LoosinCreek.NO3.monthly.ts)

summary(LoosinCreek.NO3.monthly.trend)

#subtracting seasonal component from our ts object
LoosinCreek.NO3.monthly.nonseas <- (LoosinCreek.NO3.monthly.ts - 
                                       LoosinCreek.NO3.monthly.decomposed$time.series[,1])

#making sure it worked
plot(LoosinCreek.NO3.monthly.nonseas)
plot(LoosinCreek.NO3.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
LoosinCreek.NO3.monthly.nonseas.trend <- trend::mk.test(LoosinCreek.NO3.monthly.nonseas)
# Inspect results
LoosinCreek.NO3.monthly.nonseas.trend
```
Seaonsal trend is significantly decreasing ? 
```{r}
#Temp - Loosin Creek
LoosinCreek_Temp_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Loosin Creek") %>% 
  select(Date, Temp, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
LoosinCreek_Temp_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(LoosinCreek_Temp_months) <- "Date"

#Left joining to point out gaps
LoosinCreek_Temp_timeseries_ready <- left_join(LoosinCreek_Temp_months, 
                                               LoosinCreek_Temp_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in Temp
LoosinCreek_Temp_timeseries_ready$Temp = 
  zoo::na.approx(LoosinCreek_Temp_timeseries_ready$Temp)

#Creating time series for Loosin Creek's Temp
LoosinCreek.Temp.monthly.ts <- ts(LoosinCreek_Temp_timeseries_ready$Temp, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
LoosinCreek.Temp.monthly.decomposed <- stl(LoosinCreek.Temp.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(LoosinCreek.Temp.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
LoosinCreek.Temp.monthly.trend <- 
  Kendall::SeasonalMannKendall(LoosinCreek.Temp.monthly.ts)

summary(LoosinCreek.Temp.monthly.trend)

#subtracting seasonal component from our ts object
LoosinCreek.Temp.monthly.nonseas <- (LoosinCreek.Temp.monthly.ts - 
                                       LoosinCreek.Temp.monthly.decomposed$time.series[,1])

#making sure it worked
plot(LoosinCreek.Temp.monthly.nonseas)
plot(LoosinCreek.Temp.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
LoosinCreek.Temp.monthly.nonseas.trend <- trend::mk.test(LoosinCreek.Temp.monthly.nonseas)
# Inspect results
LoosinCreek.Temp.monthly.nonseas.trend
```
Significantly increasing. 

For Loosin's Creek, NH4 showed a significant downward trend and temperature showed a significant upward trend.

Next, let's do Research Creek. 
```{r}
#PO4 - Research Creek
ResearchCreek_PO4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Research Creek") %>% 
  select(Date, PO4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ResearchCreek_PO4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ResearchCreek_PO4_months) <- "Date"

#Left joining to point out gaps
ResearchCreek_PO4_timeseries_ready <- left_join(ResearchCreek_PO4_months, 
                                               ResearchCreek_PO4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in PO4
ResearchCreek_PO4_timeseries_ready$PO4 = 
  zoo::na.approx(ResearchCreek_PO4_timeseries_ready$PO4)

#Creating time series for Research Creek's PO4
ResearchCreek.PO4.monthly.ts <- ts(ResearchCreek_PO4_timeseries_ready$PO4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ResearchCreek.PO4.monthly.decomposed <- stl(ResearchCreek.PO4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ResearchCreek.PO4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ResearchCreek.PO4.monthly.trend <- 
  Kendall::SeasonalMannKendall(ResearchCreek.PO4.monthly.ts)

summary(ResearchCreek.PO4.monthly.trend)

#subtracting seasonal component from our ts object
ResearchCreek.PO4.monthly.nonseas <- (ResearchCreek.PO4.monthly.ts - 
                                       ResearchCreek.PO4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ResearchCreek.PO4.monthly.nonseas)
plot(ResearchCreek.PO4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ResearchCreek.PO4.monthly.nonseas.trend <- trend::mk.test(ResearchCreek.PO4.monthly.nonseas)
# Inspect results
ResearchCreek.PO4.monthly.nonseas.trend
```
Insignificant trend. 
```{r}
#NH4 - Research Creek
ResearchCreek_NH4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Research Creek") %>% 
  select(Date, NH4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ResearchCreek_NH4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ResearchCreek_NH4_months) <- "Date"

#Left joining to point out gaps
ResearchCreek_NH4_timeseries_ready <- left_join(ResearchCreek_NH4_months, 
                                               ResearchCreek_NH4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in NH4
ResearchCreek_NH4_timeseries_ready$NH4 = 
  zoo::na.approx(ResearchCreek_NH4_timeseries_ready$NH4)

#Creating time series for Research Creek's NH4
ResearchCreek.NH4.monthly.ts <- ts(ResearchCreek_NH4_timeseries_ready$NH4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ResearchCreek.NH4.monthly.decomposed <- stl(ResearchCreek.NH4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ResearchCreek.NH4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ResearchCreek.NH4.monthly.trend <- 
  Kendall::SeasonalMannKendall(ResearchCreek.NH4.monthly.ts)

summary(ResearchCreek.NH4.monthly.trend)

#subtracting seasonal component from our ts object
ResearchCreek.NH4.monthly.nonseas <- (ResearchCreek.NH4.monthly.ts - 
                                       ResearchCreek.NH4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ResearchCreek.NH4.monthly.nonseas)
plot(ResearchCreek.NH4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ResearchCreek.NH4.monthly.nonseas.trend <- trend::mk.test(ResearchCreek.NH4.monthly.nonseas)
# Inspect results
ResearchCreek.NH4.monthly.nonseas.trend
```
Significant downward trend. 
```{r}
#NO3 - Research Creek
ResearchCreek_NO3_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Research Creek") %>% 
  select(Date, NO3F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ResearchCreek_NO3_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ResearchCreek_NO3_months) <- "Date"

#Left joining to point out gaps
ResearchCreek_NO3_timeseries_ready <- left_join(ResearchCreek_NO3_months, 
                                               ResearchCreek_NO3_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in NO3
ResearchCreek_NO3_timeseries_ready$NO3 = 
  zoo::na.approx(ResearchCreek_NO3_timeseries_ready$NO3)

#Creating time series for Research Creek's NO3
ResearchCreek.NO3.monthly.ts <- ts(ResearchCreek_NO3_timeseries_ready$NO3, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ResearchCreek.NO3.monthly.decomposed <- stl(ResearchCreek.NO3.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ResearchCreek.NO3.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ResearchCreek.NO3.monthly.trend <- 
  Kendall::SeasonalMannKendall(ResearchCreek.NO3.monthly.ts)

summary(ResearchCreek.NO3.monthly.trend)

#subtracting seasonal component from our ts object
ResearchCreek.NO3.monthly.nonseas <- (ResearchCreek.NO3.monthly.ts - 
                                       ResearchCreek.NO3.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ResearchCreek.NO3.monthly.nonseas)
plot(ResearchCreek.NO3.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ResearchCreek.NO3.monthly.nonseas.trend <- trend::mk.test(ResearchCreek.NO3.monthly.nonseas)
# Inspect results
ResearchCreek.NO3.monthly.nonseas.trend
```
Significant postive trend ? 
```{r}
#Temp - Research Creek
ResearchCreek_Temp_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Research Creek") %>% 
  select(Date, Temp, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ResearchCreek_Temp_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ResearchCreek_Temp_months) <- "Date"

#Left joining to point out gaps
ResearchCreek_Temp_timeseries_ready <- left_join(ResearchCreek_Temp_months, 
                                               ResearchCreek_Temp_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in Temp
ResearchCreek_Temp_timeseries_ready$Temp = 
  zoo::na.approx(ResearchCreek_Temp_timeseries_ready$Temp)

#Creating time series for Research Creek's Temp
ResearchCreek.Temp.monthly.ts <- ts(ResearchCreek_Temp_timeseries_ready$Temp, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ResearchCreek.Temp.monthly.decomposed <- stl(ResearchCreek.Temp.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ResearchCreek.Temp.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ResearchCreek.Temp.monthly.trend <- 
  Kendall::SeasonalMannKendall(ResearchCreek.Temp.monthly.ts)

summary(ResearchCreek.Temp.monthly.trend)

#subtracting seasonal component from our ts object
ResearchCreek.Temp.monthly.nonseas <- (ResearchCreek.Temp.monthly.ts - 
                                       ResearchCreek.Temp.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ResearchCreek.Temp.monthly.nonseas)
plot(ResearchCreek.Temp.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ResearchCreek.Temp.monthly.nonseas.trend <- trend::mk.test(ResearchCreek.Temp.monthly.nonseas)
# Inspect results
ResearchCreek.Temp.monthly.nonseas.trend
```
Seems like there is an outlier in the temperature data (most likley some kind of error). 

For Research Creek, NH4 had a significant downward trend and NO3 had a significant postive trend ? 

Lastly, we will do Zeke's Basin. 
```{r}
#PO4 - Zeke's Basin
ZekeBasin_PO4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Zeke's Basin") %>% 
  select(Date, PO4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ZekeBasin_PO4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ZekeBasin_PO4_months) <- "Date"

#Left joining to point out gaps
ZekeBasin_PO4_timeseries_ready <- left_join(ZekeBasin_PO4_months, 
                                               ZekeBasin_PO4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in PO4
ZekeBasin_PO4_timeseries_ready$PO4 = 
  zoo::na.approx(ZekeBasin_PO4_timeseries_ready$PO4)

#Creating time series for Research Creek's PO4
ZekeBasin.PO4.monthly.ts <- ts(ZekeBasin_PO4_timeseries_ready$PO4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ZekeBasin.PO4.monthly.decomposed <- stl(ZekeBasin.PO4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ZekeBasin.PO4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ZekeBasin.PO4.monthly.trend <- 
  Kendall::SeasonalMannKendall(ZekeBasin.PO4.monthly.ts)

summary(ZekeBasin.PO4.monthly.trend)

#subtracting seasonal component from our ts object
ZekeBasin.PO4.monthly.nonseas <- (ZekeBasin.PO4.monthly.ts - 
                                       ZekeBasin.PO4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ZekeBasin.PO4.monthly.nonseas)
plot(ZekeBasin.PO4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ZekeBasin.PO4.monthly.nonseas.trend <- trend::mk.test(ZekeBasin.PO4.monthly.nonseas)
# Inspect results
ZekeBasin.PO4.monthly.nonseas.trend
```
Insignificant trend. 
```{r}
#NH4 - Zeke's Basin
ZekeBasin_NH4_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Zeke's Basin") %>% 
  select(Date, NH4F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ZekeBasin_NH4_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ZekeBasin_NH4_months) <- "Date"

#Left joining to point out gaps
ZekeBasin_NH4_timeseries_ready <- left_join(ZekeBasin_NH4_months, 
                                               ZekeBasin_NH4_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in PO4
ZekeBasin_NH4_timeseries_ready$NH4 = 
  zoo::na.approx(ZekeBasin_NH4_timeseries_ready$NH4)

#Creating time series for Research Creek's PO4
ZekeBasin.NH4.monthly.ts <- ts(ZekeBasin_NH4_timeseries_ready$NH4, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ZekeBasin.NH4.monthly.decomposed <- stl(ZekeBasin.NH4.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ZekeBasin.NH4.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ZekeBasin.NH4.monthly.trend <- 
  Kendall::SeasonalMannKendall(ZekeBasin.NH4.monthly.ts)

summary(ZekeBasin.NH4.monthly.trend)

#subtracting seasonal component from our ts object
ZekeBasin.NH4.monthly.nonseas <- (ZekeBasin.NH4.monthly.ts - 
                                       ZekeBasin.NH4.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ZekeBasin.NH4.monthly.nonseas)
plot(ZekeBasin.NH4.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ZekeBasin.NH4.monthly.nonseas.trend <- trend::mk.test(ZekeBasin.NH4.monthly.nonseas)
# Inspect results
ZekeBasin.NH4.monthly.nonseas.trend
```
Significant downward trend. 
```{r}
#NO3 - Zeke's Basin
ZekeBasin_NO3_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Zeke's Basin") %>% 
  select(Date, NO3F, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ZekeBasin_NO3_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ZekeBasin_NO3_months) <- "Date"

#Left joining to point out gaps
ZekeBasin_NO3_timeseries_ready <- left_join(ZekeBasin_NO3_months, 
                                               ZekeBasin_NO3_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in NO3
ZekeBasin_NO3_timeseries_ready$NO3 = 
  zoo::na.approx(ZekeBasin_NO3_timeseries_ready$NO3)

#Creating time series for Research Creek's NO3
ZekeBasin.NO3.monthly.ts <- ts(ZekeBasin_NO3_timeseries_ready$NO3, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ZekeBasin.NO3.monthly.decomposed <- stl(ZekeBasin.NO3.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ZekeBasin.NO3.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ZekeBasin.NO3.monthly.trend <- 
  Kendall::SeasonalMannKendall(ZekeBasin.NO3.monthly.ts)

summary(ZekeBasin.NO3.monthly.trend)

#subtracting seasonal component from our ts object
ZekeBasin.NO3.monthly.nonseas <- (ZekeBasin.NO3.monthly.ts - 
                                       ZekeBasin.NO3.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ZekeBasin.NO3.monthly.nonseas)
plot(ZekeBasin.NO3.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ZekeBasin.NO3.monthly.nonseas.trend <- trend::mk.test(ZekeBasin.NO3.monthly.nonseas)
# Inspect results
ZekeBasin.NO3.monthly.nonseas.trend
```
Insignificant trend. 
```{r}
#Temp - Zeke's Basin
ZekeBasin_Temp_monthly <- stations_wq_nutr_cleaned %>%
  filter(StationName == "Zeke's Basin") %>% 
  select(Date, Temp, Year, Month) %>% 
  filter(Date > as.Date("2002-03-01")) %>% 
  filter(Date < as.Date("2025-01-01"))

#Finding gaps by making new data frame
ZekeBasin_Temp_months <- as.data.frame(seq.Date(from=as.Date("2002/04/01"),
                                 to=as.Date("2024/12/01"),
                                 by="month"))
colnames(ZekeBasin_Temp_months) <- "Date"

#Left joining to point out gaps
ZekeBasin_Temp_timeseries_ready <- left_join(ZekeBasin_Temp_months, 
                                               ZekeBasin_Temp_monthly, 
                                               by = "Date")

#Using linear interpolation to fill in Temp
ZekeBasin_Temp_timeseries_ready$Temp = 
  zoo::na.approx(ZekeBasin_Temp_timeseries_ready$Temp)

#Creating time series for Research Creek's Temp
ZekeBasin.Temp.monthly.ts <- ts(ZekeBasin_Temp_timeseries_ready$Temp, 
                                  frequency = 12, start = c(2002,3))

#Decomposing
ZekeBasin.Temp.monthly.decomposed <- stl(ZekeBasin.Temp.monthly.ts, 
                                           s.window = "periodic")

#plotting
plot(ZekeBasin.Temp.monthly.decomposed)

#Running monotonic trend analysis to inspect seasonality
ZekeBasin.Temp.monthly.trend <- 
  Kendall::SeasonalMannKendall(ZekeBasin.Temp.monthly.ts)

summary(ZekeBasin.Temp.monthly.trend)

#subtracting seasonal component from our ts object
ZekeBasin.Temp.monthly.nonseas <- (ZekeBasin.Temp.monthly.ts - 
                                       ZekeBasin.Temp.monthly.decomposed$time.series[,1])

#making sure it worked
plot(ZekeBasin.Temp.monthly.nonseas)
plot(ZekeBasin.Temp.monthly.ts)

# Running Mann Kendall on non-seasonal monthly series
ZekeBasin.Temp.monthly.nonseas.trend <- trend::mk.test(ZekeBasin.Temp.monthly.nonseas)
# Inspect results
ZekeBasin.Temp.monthly.nonseas.trend
```
Significant upward trend. 

For Zeke's Basin, NH4 had a significant downward trend and temperature had a significant upward trend. 




## Summary and Conclusion


\newpage

## References
